Embedding intuitionistic into classical logic

### 2022-08-17 Notes/Embedding intuitionistic into classical logic/EIICL.tex
First Final Thesis Draft

\documentclass[a4paper,12pt]{report}

\title{Embedding intuitionistic into classical logic}
\author{Alexander Pluska}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{ wasysym }
\usepackage{ stmaryrd }
\usepackage{ mathpartir }
\usepackage{bussproofs}
\usepackage{enumerate}
\usepackage{tikz-qtree}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{arrows,shapes,quotes}
\usepackage{stackengine}
\usepackage{bbold}
\usepackage{footmisc}
\usepackage{wrapfig}

@@

\DeclareMathOperator*{\argmax}{arg\,max}

@@

	\maketitle

	\begin{abstract}
		We present an embedding of intuitionistic into classical logic, a quasi-inverse to the famous double-negation translation, i.e. a procedure that for each formula $\varphi$ gives a formula $\varphi_c$ such that $\varphi$ is intuitionistically valid if and only if $\varphi_c$ is classically valid. We consider both the propositional and predicate case. Since many of the more involved results for the more interesting predicate scenario involve similar arguments as in the propositional case we shall develop them simultaneously.
	\end{abstract}

	\tableofcontents

	\chapter{Preliminaries}

	First let us recapitulate the most important definitions and some facts about them. This will serve to fix our notation as well.

@@

	\section{Proof theory}

	Logical truth can also be approached via syntax rather than semantics. This gives rise to proof theory which will be especially important in the third chapter.

	\begin{definition}
		A \textit{sequent} $\Gamma\vdash \Delta$ consists of two multisets of formulas $\Gamma, \Delta$. The following rules make up our classical calculus $\mathbf{GC}$, corresponding roughly to \textbf{G3c} from~\cite{basicprooftheory}:\\
		\begin{center}
			\begin{tabular}{lll}
				\AxiomC{\hphantom{x}}
				\RightLabel{Ax ($P$ atomic)}
				\UnaryInfC{$P,\Gamma\vdash \Delta, P$}
				\DisplayProof&
				\AxiomC{\hphantom{x}}
				\RightLabel{L$\bot$}
				\UnaryInfC{$\bot,\Gamma\vdash\Delta$}
				\DisplayProof&
				\\&&\\
				\AxiomC{$A, B,\Gamma\vdash\Delta$}
				\RightLabel{L$\wedge$}
				\UnaryInfC{$A\wedge B, \Gamma\vdash \Delta$}
				\DisplayProof&
				\AxiomC{$\Gamma\vdash\Delta, A$}
				\AxiomC{$\Gamma\vdash\Delta, B$}
				\RightLabel{R$\wedge$}
				\BinaryInfC{$\Gamma\vdash \Delta, A\wedge B$}
				\DisplayProof&
				\\&&\\
				\AxiomC{$A, \Gamma\vdash\Delta$}
				\AxiomC{$B, \Gamma\vdash\Delta$}
				\RightLabel{L$\vee$}
				\BinaryInfC{$A\vee B, \Gamma\vdash \Delta$}
				\DisplayProof&
				\AxiomC{$\Gamma\vdash\Delta, A, B$}
				\RightLabel{R$\vee$}
				\UnaryInfC{$\Gamma\vdash \Delta, A\vee B$}
				\DisplayProof&
				\\&&\\
				\AxiomC{$\Gamma\vdash\Delta, A$}
				\AxiomC{$B, \Gamma\vdash\Delta$}
				\RightLabel{L$\to$}
				\BinaryInfC{$A\to B, \Gamma\vdash \Delta$}
				\DisplayProof&
				\AxiomC{$A,\Gamma\vdash\Delta, B$}
				\RightLabel{R$\to$}
				\UnaryInfC{$\Gamma\vdash \Delta, A\to B$}
				\DisplayProof&
				\\&&\\
				\AxiomC{$A[t/x], \Gamma\vdash\Delta$}
				\RightLabel{L$\forall$}
				\UnaryInfC{$\forall xA, \Gamma\vdash \Delta$}
				\DisplayProof&
				\AxiomC{$\Gamma\vdash\Delta, A[a/x]$}
				\RightLabel{R$\forall$}
				\UnaryInfC{$\Gamma\vdash \Delta, \forall xA$}
				\DisplayProof&
				\\&&\\
				\AxiomC{$A[a/x], \Gamma\vdash\Delta$}
				\RightLabel{L$\exists$}
				\UnaryInfC{$\exists xA, \Gamma\vdash \Delta$}
				\DisplayProof&
				\AxiomC{$\Gamma\vdash\Delta, A[t/x]$}
				\RightLabel{R$\exists$}
				\UnaryInfC{$\Gamma\vdash \Delta, \exists xA$}
				\DisplayProof&
				\\&&\\
				\AxiomC{$t = t, \Gamma\vdash \Delta$}
				\RightLabel{Ref}
				\UnaryInfC{$\Gamma\vdash \Delta$}
				\DisplayProof&
				\AxiomC{$s = t, P[s/x], P[t/x], \Gamma\vdash \Delta$}
				\RightLabel{Ref}
				\UnaryInfC{$s = t, P[s/x],\Gamma\vdash \Delta$}
				\DisplayProof&
				\\&&\\
			\end{tabular}
		\end{center}

		where in R$\forall$ and L$\exists$ $a$ is a free variable not occurring in the sequent otherwise.
		The above rules make up the calculus called \textbf{Gc}. It corresponds to \textbf{G3c}$^=$ from~\cite{basicprooftheory}.

		For intuitionistic logic we remain with a multi-succedent system by adjusting R$\to$ and R$\forall$:\\

		\begin{center}
			\AxiomC{$A, \Gamma\vdash B$}
			\RightLabel{R$\to$}
			\UnaryInfC{$\Gamma\vdash\Delta, A\to B$}
			\DisplayProof\hspace{2cm}
			\AxiomC{$\Gamma\vdash A[a/x]$}
			\RightLabel{R$\forall$}
			\UnaryInfC{$\Gamma\vdash\Delta, \forall xA$}
			\DisplayProof
		\end{center}

		We will call the calculus containing this formulas $\mathbf{GI}$ roughly corresponding to $\mathbf{G3i}$ from~\cite{basicprooftheory}. Usually for intuitionistic logic a single-formula succedent system is considered, but we are interested in reducing the difference between the calculi as much as possible.

		A \textit{proof} of a formula $\varphi$ in a calculus $\mathbf{G}$ is a labelled binary rooted tree in which every node is labelled with a sequent such that
		\begin{itemize}
			\item Every leaf is labelled in accordance to Ax or L$\bot$.
			\item The sequent of an interior node can be derived from sequents of its child nodes according to one of the rules.
		\end{itemize}
		We usually present a proof directly as a sequence of applications of rules rather than in a more usual tree presentation. If a sequent $\mathcal S\vdash\mathcal T$ is provable in $\mathbf{GC}$ we write $\mathcal S\vdash_C\mathcal T$, if it is in $\mathbf{GI}$ we write $\mathcal S\vdash_I\mathcal T$.
	\end{definition}

	\begin{example}
		The following is a proof of $(A\to B)\to A\vdash B$:
		\begin{center}
			\AxiomC{}
			\RightLabel{Ax}
			\UnaryInfC{$A\to B, A\vdash B, A$}
			\AxiomC{}
			\RightLabel{Ax}
			\UnaryInfC{$A, B\vdash B$}
			\RightLabel{L$\to$}
			\BinaryInfC{$A\to B, A\vdash B$}
			\DisplayProof
		\end{center}
	\end{example}

	\noindent We now recall some important properties of the given systems:

	\begin{theorem}[subformula property]
		Any formula that appears in a derivation is a subformula of a forumla that occurs in the root sequent.
	\end{theorem}

>>>>

\documentclass[a4paper,11pt]{report}

\usepackage{setspace}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{ wasysym }
\usepackage{ stmaryrd }
\usepackage{ mathpartir }
\usepackage{bussproofs}
\usepackage{enumerate}
\usepackage{tikz-qtree}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{arrows,shapes,quotes}
\usepackage{stackengine}
\usepackage{footmisc}
\usepackage{wrapfig}
\usepackage[a4paper,top=2.5cm,bottom=3.5cm,left=2.5cm,right=2.5cm,headsep=10pt]{geometry}
\usepackage[toc,page]{appendix}
\usepackage{listings}
\lstset{
	language=bash,
	basicstyle=\ttfamily
}

@@

@@

	\pagenumbering{gobble}
	\begin{flushright}
		% https://public.univie.ac.at/en/downloads/logos/
		\includegraphics[width=7cm]{RZ_Logo_Uni_sw}
	\end{flushright}	
	\vskip20mm
	\begin{center}
		\huge\textbf{MASTERARBEIT / MASTER'S THESIS}
		\vskip15mm
		\normalsize{Titel der Masterarbeit / Title of the Master's Thesis}

		\LARGE\textbf{``Automated Deduction for Intuitionistic Logic\\via Embedding into Classical Logic''} % Insert title here
		\vskip15mm
		\normalsize{verfasst von / submitted by}

		\Large\textbf{Alexander Pluska, B.Sc.} % Insert your name in this line
		\vskip15mm
		\normalsize{angestrebter akademischer Grad / in partial fulfillment of the requirements for the degree of}

		\Large\textbf{Master of Science (M.Sc.)}

		\small
	\end{center}
	\vskip15mm
	\begingroup
	\renewcommand{\arraystretch}{1.2}
	\begin{flushleft}
		Wien, 2022 / Vienna 2022 % Insert graduation year in this line
		\vfill
		\addtolength{\tabcolsep}{-6pt} 
		\begin{table}[h!]
			\begingroup
			\renewcommand*{\arraystretch}{1.2}
			%\renewcommand*{\linespacing}{1}
			\renewcommand\baselinestretch{1}
			\small
			\begin{tabular}{p{7cm} p{10cm}}
				Studienkennzahl lt. Studienblatt / \newline degree programme code as it appears on \newline the student record sheet: &A 066 821 \\
				Studienrichtung lt. Studienblatt / \newline degree programme as it appears on \newline the student record sheet:& Mathematics \\
				Betreut von / Supervisor: & Univ.-Prof. Matthias Aschenbrenner \\ % Insert your program/supervisor here
				Mitbetreut von / Co-Supervisor: & Ass.-Prof. Florian Zuleger \\ 
			\end{tabular}
			\endgroup
		\end{table}
		\addtolength{\tabcolsep}{6pt} 
	\end{flushleft}	
	\endgroup

	\newpage
	\
	\newpage

	% Some of the code for creating ToC id adapted from the following template:
	% https://www.sharelatex.com/templates/thesis/university-of-sussex-thesis

		\
		\vfill
		\onehalfspacing
		\begin{center}
		\bfseries Abstract		
		\end{center}
		The famous double negation translation~\cite{glivenko1929quelques, godel1933intuitionistischen} establishes an embedding from classical into intuitionistic logic. Curiously the reverse direction has not been covered in literature. We present an effective embedding from intuitionistic into classical logic, both in the propositional and first-order case, as well as an effective embedding of intuitionistic propositional logic into quantified boolean formulas.

		Furthermore we implement a practical system that takes intuitionistic first-order problems as an input and performs our transformation. This allows the use of classical theorem provers for establishing intuitionistic validity. We benchmark our implementation using the vampire theorem prover~\cite{kovacs2013first} on the ILTP problem set~\cite{iltp}.
		%The key notion used in the proofs is the existence of counter-models - an analogon to the satisfiability of the negation in classical logic, which is not equivalent to invalidity in the intuitionistic case.  We hope that this will allow leveraging the tremendous recent progress in automated reasoning in classical logic for checking intuitionistic validity, for which there has not been much progress in recent decades.

		Finally we discuss how the generated classical proofs of the transformed problems can be translated to intuitionistic proofs of the original problem. All in all this establishes a novel approach to theorem proving for intuitionistic logic.

	\vspace*{1cm}
	\begin{center}
		\bfseries Acknowledgements		
	\end{center}

	I want to thank my Matthias for accommodating my particular interests by agreeing to supervise a thesis which lies outside of his main area of research and bearing with the organizational difficulties. I want to thank Florian who graciously agreed to co-supervise a thesis at a different university. He has provided me with a fascinating topic as well as plenty of his time, expertise and thorough feedback. Finally I want to thank my family, in particular my parents Stefania and Cezary, who have instilled in me the joy of learning, provided all the love and help I could ask for and invariably supported me along my journey.

	{\vfill\null}

	\newpage
	\
	\singlespacing
	\newgeometry{top=1cm}
	\tableofcontents
	\restoregeometry

	\chapter{Introduction}
	\onehalfspacing

	\pagenumbering{arabic}
	Constructive mathematics refers to a flavour of mathematics in which the existence of an object can only be established by explicit construction, as opposed to classical mathematics where existence can be shown implicitly, e.g. by assuming non-existence and deriving a contradiction.
	The formalism usually associated with constructive mathematics is intuitionistic logic, which essentially differentiates itself from classical logic by the fact that the law of excluded middle $A\vee\neg A$ and the double negation shift $\forall x\neg\neg P(x)\to\neg\neg\forall xP(x)$ are not valid.
	Besides philosophical considerations, most prominently advocated by Brouwer~\cite{brouwer1907over} and Bishop~\cite{bishop1967foundations}, there is a particular motivation for studying constructive mathematics from the perspective of computer science in that proofs in  intuitionistic logic directly correspond to computer programs --- as expressed in the Curry-Howard correspondence~\cite{howard1980formulae}.

	The interest in intuitionistic logic has lead to the development of a number of automated theorem proving systems for propositional as well as for predicate logic and a collection of benchmark problems (see e.g. the
	ILTP library website~\cite{iltp}).
	However, progress in automated theorem proving for intuitionistic logic has been slow, whereas reasoners for classical logic have made tremendous progress, see e.g. the TPTP~\cite{casc} and SAT~\cite{satc} competitions.
	%, e.g., the first-order provers~\cite{kovacs2013first, schulz2002brainiac, korovin2008iprover}
	%Comparing benchmarks for classical and intuitionistic automated theorem proving, e.g. TPTP~\cite{casc} and , we see that classical provers currently have a far better success rate. Furthermore all the most popular first-order utilize classical logic.
	This difference can be partially explained by fundamental differences between the logics.
	First of all, determining intuitionistic validity is computationally harder, i.e. in the propositional case intuitionistic validity is \verb+PSPACE+-complete~\cite{statman1979intuitionistic} whereas classical validity is \verb+coNP+-complete~\cite{cook1971complexity}.
	A further advantage of classical logic is the existence of calculi that are particularly suitable for automation such as superposition~\cite{bachmair2001resolution}, which rely on the existence of convenient normal forms such as CNF, and the duality between validity and satisfiability, i.e. to show the validity of a formula it suffices to show the unsatisfiability of the negated formula.
	While some (albeit more complex) normal forms also exist  for intuitionistic logic, crucially the duality between validity and satisfiability of the negation does not hold.
	Therefore most dedicated intuitionistic theorem provers~\cite{mclaughlin2009efficient, tammet1996resolution} use the naive inverse method, i.e. directly search for a cut-free proof by applying the rules from some proof calculus inversely, enhanced by search strategies such as focussing and polarization. This approach generally leads to a much more complex search and is therefore difficult to implement efficiently.
	Finally, we add that in contrast to intuitionistic provers a tremendous amount of work has been put into optimizing provers for classical logic, in particular for the propositional case, i.e. SAT-solvers.

	With this work we want to propose a novel approach for intuitionistic theorem proving that leverages the progress in classical theorem proving, that is for each formula $\varphi$:
	\begin{itemize}
		\item Give a formula $\varphi^\#$ that is classically valid if and only if $\varphi$ is intuitionistically valid.
		\item Use a state-of-the-art classical prover to establish the validity/invalidity of $\varphi^\#$.
		\item Transform the output proof/counter-model of $\varphi^\#$ to one of $\varphi$.
	\end{itemize}
	The most challenging part of this approach is giving the translation of $\varphi$ to $\varphi^\#$.
	Interestingly, the reverse direction, the famous double-negation translation, has long been established and goes back to Glivenko~\cite{glivenko1929quelques} in the propositional case, and to G\"odel~\cite{godel1933intuitionistischen} and Gentzen~\cite{gentzen1936widerspruchsfreiheit} in the first-order case. In the propositional case it is particularly simple: $\varphi$ is classically valid if and only if $\neg\neg\varphi$ is intuitionistically valid. Intuitively the translation collapses for each subformula $\psi$ of $\varphi$ the truth values of $\psi$ and $\neg\neg\psi$, which are classically but not intuitionistically equivalent. This gives us a first idea why the reverse direction is perhaps more difficult: We need to expand the truth values of $\psi$ and $\neg\neg\psi$, i.e. if they both occur in $\varphi$, we must have a way to (classically) assign different truth values to their respective counterparts in $\varphi^\#$. In particular, this necessitates the introduction of new propositional variables in our procedure, which already marks a big difference to the double negation translation.
	%
	While establishing the translation of formulas we also perform an effective translation of counter-models, i.e. for each intuitionistic counter-model of $\varphi$ we effectively construct a classical counter-model of $\varphi^\#$ and vice versa.
	We note that the existence of counter-models is a key notion that forms a proper dual to validity --- whereas the satisfiability of the negation is not necessary for invalidity (in contrast to classical logic, where it is a proper dual to validity).
	Transforming and reducing counter-models to a normal form is also what ultimately enables our translation. Apart from a translation of counter-models we also explicitly describe how to equivalently transform classical proofs of $\varphi^\#$ to intuitionistic proofs of $\varphi$, in particular resolution proofs as given by theorem provers like vampire~\cite{Kov_cs_2013}.

	As a final contribution we implement our translation in the Rust programming language. It takes a first-order problem in the tptp format~\cite{tptp} and outputs the translated problem in the same format. The code is published on github~\cite{implementation}. We then benchmark our implementation using the ILTP problem set~\cite{iltp}, i.e. we translate all problems in the set and then run the vampire theorem prover~\cite{Kov_cs_2013} on the translated problems. Our approach performs on par with most existing approaches for intuitionistic theorem provers, but comes short of the state-of-the art leaders. As there is still a lot of room for optimization this is a hopeful sign.

	\chapter{Overview}

	The main content is organized in chapters 3-6. Chapter 3 collects prerequisite definitions and serves to fix our notation and recall some important facts. Since this overview already contains some technical results it may be consulted for reference. Chapter 4 is the meat of this thesis. In particular it features a complete description of our embedding of intuitionistic into classical logic as well as proving its correctness. In chapter 5 we present our implementation and benchmark it. Finally in chapter 6 we explicitly describe how to translate the generated proofs and counter-models which in principle directly follows from the results of section 4. 	We shall now for each chapter give an overview of our main results and highlight the key arguments.

	\section{Embedding Intuitionistic into Classical Logic}

	Recall that our goal in this chapter is to give an effective translation procedure that, for a given formula $\varphi$, yields a formula $\varphi^\#$, such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid. We start with the propositional case. The key arguments are the same as in the first-order case, while it is technically simpler and less cluttered.

	Before our main transformation we employ a preprocessing step akin to the Tseytin transformation~\cite{tseitin1983complexity}, which is a popular pre-processing step in classical automated reasoning:
	It gives an equisatisfiable sentence (over an extended set of propositions) in conjunctive normal form.
	Eliminating all implications, however, is not possible in intuitionistic logic since $A\to B$ is not equivalent to $\neg A\vee B$.
	Still we propose a similar transformation.
	A notable feature of our transformation is that all non-classical content is encapsulated in formulas of type $(A\to B)\to C$, i.e. if there are no such formulas in the transformed formula then the classical validity of the transformed formula immediately implies intuitionistic validity.

	The main transformation then proceeds in three steps:
	%
	1) We encode as a first-order sentence that the considered formula holds for every Kripke frame.
	Since Kripke frames for propositional logic over a fixed alphabet of propositional variables form a first-order theory this step is rather straightforward.
	%
	2) Next we eliminate some quantifiers via Herbrandization. The only quantifiers eliminated occur in formulas derived from formulas of type $(A\to B)\to C$.
	%
	3) We can then argue that if there is a counter-model for the resulting formula then there is a counter-model with a certain Kripke frame of bounded size completely determined by the number of formulas of type $(A\to B)\to C$.
	This allows us to eliminate the remaining quantifiers by simply enumerating the worlds in that Kripke frame.

	Summarizing, we obtain the following result:

	\begin{theorem}
		\label{thm:reduction-propositional}
		Let $\mathcal S$ be as in Theorem~\ref{thm:Tseytin} and $\mathcal F_\to\subseteq\mathcal S$ denote the subset of formulas of the form $(A\to B)\to C$ and $\Lambda$ denote the set of sequences without repetition over $\mathcal F_\to$. For each atom $A$ and $k\in\Lambda$ consider a new atom $A^k$. Obtain $\mathcal S^\#$ by including the following formulas:
		\begin{itemize}
			\item $A^k\to A^{k\psi}$ for each atom $A$ occurring in $\mathcal S$, $k\in\Lambda$ and $\psi\in\mathcal F_\to$ not occurring in $k$.
			\item $A^k\to (B^k\circ C^k)$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
			\item $(A^k\circ B^k)\to C^k$ for each $\circ\in\{\wedge,\vee\}$, $A\to (B\circ C)\in\mathcal S$, $k\in\Lambda$.
			\item $(A^{k\psi}\to B^{k\psi})\to C^k$ for $\psi = (A\to B)\to C\in\mathcal S$, $k\in\Lambda$ if $\psi$ does not occurr in $k$.
		\end{itemize}
		Then, $\bigwedge S\to P$ is intuitionistically valid iff $\bigwedge S^\#\to P^\epsilon$ is classically valid, where $\epsilon$ denotes the empty sequence. The size of $\mathcal S^\#$ is in $\mathcal O(|\mathcal S|\cdot2^{|\mathcal F_\to|\cdot\log(|\mathcal F_\to|)})$. Futhermore, there is an effective procedure for translating counter-models between the sentences.
	\end{theorem}

	Instead of explicitly enumerating the worlds of a counter-example of bounded size (as in the reduction stated in the above theorem), we can also do this enumeration implicitly with a quantified boolean formula (QBF).
	We can show that this QBF is satisfiable if and only if there exists a counter-model for our original formula:

	\begin{theorem}
		Let $\mathcal S$ be as in Theorem~\ref{thm:Tseytin} and $\mathcal F_\to\subseteq\mathcal S$ denote the subset of formulas of the form $(A\to B)\to C$. There is an effective procedure that produces a QBF $\varphi^Q$ of size $\mathcal O(|\mathcal S|\cdot|\mathcal F_\to| + |\mathcal F_\to|^3)$ with $2\cdot |\mathcal F_\to|-1$ quantifier alternations such that $\varphi$ is intuitionistically valid if and only if $\varphi^Q$ is not satisfiable. For fixed $N\in\mathbb N$, deciding intuitionistic validity for formulas $\bigwedge \mathcal S\to P$ where $\mathcal S$ is as above and $|\mathcal F_\to| = N$ is therefore in $\Sigma_{2N-1}$.
	\end{theorem}

	We then move to the first-order case. Here the construction is a bit more involved but the underlying approach is the same.
	Again we first perform a Tseytin-like transformation.
	Here the non-classical content is encapsulated by formulas of type $(A\to B)\to C$ and $\forall xA\to B$.
	The details of this translation are given in Section~\ref{section:tseytin}.
	We then proceed similarly to the propositional case.
	Note however that one difficulty arises from the fact that our (classical) domain will now contain on the one hand worlds in the Kripke frame but on the other hand also proper domain elements.
	We resolve this apparent conflict by introducing a special binary predicate $E$, inspired by~\cite{iemhoff2010eskolemization}, encoding which domain elements exists at which world.
	The main transformation then again proceeds in three steps:
	Step 1) and 2) are analogous to the propositional case.
	The real difference arises in step 3):
	While we can also establish the existence of a canonic counter-model whose frame only depends on formulas of type $(A\to B)\to C$ and $\forall xA\to B$, the size of this counter-model is countably infinite in general.
	This is not surprising since certain intuitionistically invalid formulas like the double negation shift $\forall x\neg\neg A(x)\to \neg\neg\forall x A(x)$ don't have finite counter-models.
	Therefore we are not able to eliminate the $\forall$-quantifiers associated with the Kripke semantics.
	However since our translation targets first-order logic, we believe that the introduction of these quantifiers is acceptable.
	%
	Summarizing, we obtain the following result:

	\begin{theorem}
		\label{thm:reduction-first-order-short}
		There exists a linear-time procedure that gives for every first-order formula $\varphi$ a formula $\varphi^\#$ such that $\varphi$ is intuitionistically valid if and only if $\varphi^\#$ is classically valid. The size of $\varphi^\#$ is in linear in the size of $\varphi$, however, for each $n$-ary relation symbol in $\varphi$ there is a corresponding $n+1$-ary relation symbol in $\varphi^\#$, and $\varphi$ contains a new binary predicate $E$ as well as a number of new function symbols. Furthermore, there is an effective translation between intuitionistic counter-models of $\varphi$ and classical counter-models of $\varphi^\#$.
	\end{theorem}
	A more detailed account of Theorem~\ref{thm:reduction-first-order-short} is given in Theorem~\ref{fullFOtranslation}.

	\section{Implementation}

	In this section we describe our system that implements the translation from the previous section. The rust source code can be found online~\cite{implementation}. We then elaborate our benchmarking process which utilizes the ILTP problem set~\cite{iltp} and the vampire theorem prover~\cite{Kov_cs_2013}.

	\begin{figure}[h]\label{results}
		$$\begin{matrix}
			&\text{AGT}&\text{ALG}&\text{COM}&\text{CSR}&\text{GEO}&\text{GRA}&\text{GRP}&\text{HAL}\\
			\text{Proven}&5&170&0&0&58&3&4&2\\
			\text{Disproven}&0&0&0&0&0&0&0&\\
			\text{Timeout}&47&29&3&29&107&74&1&16
		\end{matrix}$$
		$$\begin{matrix}
			&\text{KRS}&\text{LCL}&\text{MGT}&\text{MSC}&\text{NLP}&\text{NUM}&\text{PLA}&\text{PUZ}\\
			\text{Proven}&0&0&30&1&9&1&9&2\\
			\text{Disproven}&0&0&0&0&12&0&0&0\\
			\text{Timeout}&5&9&128&3&69&1&237&82
		\end{matrix}$$
		$$\begin{matrix}
			&\text{SET}&\text{SWC}&\text{SWV}&\text{SYN}&\text{TOP}&\text{GEJ}&\text{GPJ}&\text{SYJ}\\
			\text{Proven}&0&5&27&1&64&65&54&1\\
			\text{Disproven}&6&1&297&423&154&108&298&2
		\end{matrix}$$
		\caption{Performance on problem sections of the ILTP set.}
	\end{figure}

	Our implementation performs similar to existing dedicated provers for intuitionistic logic, but falls short of the state-of-the-art. Since there is still a lot of possible optimizations left for our translation this is a hopeful first sign.

	\section{Translation of proofs and counter-models}

	In the final section we give an explicit mechanism how to transform the proofs and counter-models for translated formula $\varphi^\#$ into intuitionistic proofs/counter-models of the original formula $\varphi$. The existence of these transformations is clear from the methodology of chapter~\ref{chapter:eiicl}, however they are hidden in the proofs and here we make them explicit. The two main results are as follows:
	\begin{theorem}
		There exists an effective linear time procedure that converts any classical counter-model for a translated formula $\varphi^\#$ into a intuitionistic counter-model for the original formula $\varphi$.
	\end{theorem}
	\begin{theorem}
		There exists an effective linear time procedure that converts any classical proof of a translated formula $\varphi^\#$ into a intuitionistic proof for the original formula $\varphi$.
	\end{theorem}

	Given the arguments of Chapter~\ref{chapter:eiicl} the translation of counter-models is more or less just a compilation of proofs. On the other hand the explicit translation of proofs does require some work, since to that point all arguments were semantic and this requires a syntactic viewpoint.

	\chapter{Preliminaries}
	First let us recapitulate the most important definitions and some facts about them. This will serve to fix our notation as well.

@@

	\section{Proof theory}\label{proof-theory}

	Logical truth can also be approached via syntax rather than semantics. This gives rise to proof theory which will be especially important in the final chapter when we establish proof translations.

	\begin{definition}
		Pairs of multisets of formulas $A = \{A_1,\dots, A_n\}, B = \{B_1,\dots, B_m\}$ form a \textit{Sequent} $A\Rightarrow B$. Usually we will write $A_1,\dots, A_n\Rightarrow B_1,\dots, B_m$, note however that $A$ and $B$ are \textbf{not} to be interpreted as sequences but as multisets.
	\end{definition}

	We define sequent calculi as in~\cite[p.77]{basicprooftheory}. Have following inferences\\

	\begin{center}
	\begin{tabular}{lll}
		\AxiomC{\hphantom{x}}
		\RightLabel{Ax ($P$ atomic)}
		\UnaryInfC{$P,\Gamma\Rightarrow \Delta, P$}
		\DisplayProof&
		\AxiomC{\hphantom{x}}
		\RightLabel{L$\bot$}
		\UnaryInfC{$\bot,\Gamma\Rightarrow\Delta$}
		\DisplayProof&
		\\&&\\
		\AxiomC{$A, B,\Gamma\Rightarrow\Delta$}
		\RightLabel{L$\wedge$}
		\UnaryInfC{$A\wedge B, \Gamma\Rightarrow \Delta$}
		\DisplayProof&
		\AxiomC{$\Gamma\Rightarrow\Delta, A$}
		\AxiomC{$\Gamma\Rightarrow\Delta, B$}
		\RightLabel{R$\wedge$}
		\BinaryInfC{$\Gamma\Rightarrow \Delta, A\wedge B$}
		\DisplayProof&
		\\&&\\
		\AxiomC{$A, \Gamma\Rightarrow\Delta$}
		\AxiomC{$B, \Gamma\Rightarrow\Delta$}
		\RightLabel{L$\vee$}
		\BinaryInfC{$A\vee B, \Gamma\Rightarrow \Delta$}
		\DisplayProof&
		\AxiomC{$\Gamma\Rightarrow\Delta, A, B$}
		\RightLabel{R$\vee$}
		\UnaryInfC{$\Gamma\Rightarrow \Delta, A\vee B$}
		\DisplayProof&
		\\&&\\
		\AxiomC{$\Gamma\Rightarrow\Delta, A$}
		\AxiomC{$B, \Gamma\Rightarrow\Delta$}
		\RightLabel{L$\to$}
		\BinaryInfC{$A\to B, \Gamma\Rightarrow \Delta$}
		\DisplayProof&
		\AxiomC{$A,\Gamma\Rightarrow\Delta, B$}
		\RightLabel{R$\to$}
		\UnaryInfC{$\Gamma\Rightarrow \Delta, A\to B$}
		\DisplayProof&
		\\&&\\
		\AxiomC{$A[t/x], \Gamma\Rightarrow\Delta$}
		\RightLabel{L$\forall$}
		\UnaryInfC{$\forall xA, \Gamma\Rightarrow \Delta$}
		\DisplayProof&
		\AxiomC{$\Gamma\Rightarrow\Delta, A[a/x]$}
		\RightLabel{R$\forall$}
		\UnaryInfC{$\Gamma\Rightarrow \Delta, \forall xA$}
		\DisplayProof&
		\\&&\\
		\AxiomC{$A[a/x], \Gamma\Rightarrow\Delta$}
		\RightLabel{L$\exists$}
		\UnaryInfC{$\exists xA, \Gamma\Rightarrow \Delta$}
		\DisplayProof&
		\AxiomC{$\Gamma\Rightarrow\Delta, A[t/x]$}
		\RightLabel{R$\exists$}
		\UnaryInfC{$\Gamma\Rightarrow \Delta, \exists xA$}
		\DisplayProof&
		\\&&\\
	\end{tabular}
	\end{center}
	where in R$\forall$ and L$\exists$ $a$ is a free variable not occurring in the sequent otherwise.

	The above rules make up the calculus for classical logic called \textbf{Gc}. It corresponds to \textbf{G3c} from~\cite{basicprooftheory}.
	Similarly we can define a calculus \textbf{Gi} for intuitionistic logic corresponding to \textbf{m-G3i}:\\

		\begin{center}
		\begin{tabular}{lll}
			\AxiomC{\hphantom{x}}
			\RightLabel{Ax ($P$ atomic)}
			\UnaryInfC{$P,\Gamma\Rightarrow \Delta, P$}
			\DisplayProof&
			\AxiomC{\hphantom{x}}
			\RightLabel{L$\bot$}
			\UnaryInfC{$\bot,\Gamma\Rightarrow\Delta$}
			\DisplayProof&
			\\&&\\
			\AxiomC{$A, B,\Gamma\Rightarrow\Delta$}
			\RightLabel{L$\wedge$}
			\UnaryInfC{$A\wedge B, \Gamma\Rightarrow \Delta$}
			\DisplayProof&
			\AxiomC{$\Gamma\Rightarrow\Delta, A$}
			\AxiomC{$\Gamma\Rightarrow\Delta, B$}
			\RightLabel{R$\wedge$}
			\BinaryInfC{$\Gamma\Rightarrow \Delta, A\wedge B$}
			\DisplayProof&
			\\&&\\
			\AxiomC{$A, \Gamma\Rightarrow\Delta$}
			\AxiomC{$B, \Gamma\Rightarrow\Delta$}
			\RightLabel{L$\vee$}
			\BinaryInfC{$A\vee B, \Gamma\Rightarrow \Delta$}
			\DisplayProof&
			\AxiomC{$\Gamma\Rightarrow\Delta, A, B$}
			\RightLabel{R$\vee$}
			\UnaryInfC{$\Gamma\Rightarrow \Delta, A\vee B$}
			\DisplayProof&
			\\&&\\
			\AxiomC{$A\to B, \Gamma\Rightarrow\Delta, A$}
			\AxiomC{$B, \Gamma\Rightarrow\Delta$}
			\RightLabel{L$\to$}
			\BinaryInfC{$A\to B, \Gamma\Rightarrow \Delta$}
			\DisplayProof&
			\AxiomC{$A,\Gamma\Rightarrow B$}
			\RightLabel{R$\to$}
			\UnaryInfC{$\Gamma\Rightarrow \Delta, A\to B$}
			\DisplayProof&
			\\&&\\
			\AxiomC{$A[t/x], \Gamma\Rightarrow\Delta$}
			\RightLabel{L$\forall$}
			\UnaryInfC{$\forall xA, \Gamma\Rightarrow \Delta$}
			\DisplayProof&
			\AxiomC{$\Gamma\Rightarrow A[a/x]$}
			\RightLabel{R$\forall$}
			\UnaryInfC{$\Gamma\Rightarrow \forall xA$}
			\DisplayProof&
			\\&&\\
			\AxiomC{$A[a/x], \Gamma\Rightarrow\Delta$}
			\RightLabel{L$\exists$}
			\UnaryInfC{$\exists xA, \Gamma\Rightarrow \Delta$}
			\DisplayProof&
			\AxiomC{$\Gamma\Rightarrow\Delta, A[t/x]$}
			\RightLabel{R$\exists$}
			\UnaryInfC{$\Gamma\Rightarrow \Delta, \exists xA$}
			\DisplayProof&
			\\&&\\
		\end{tabular}
	\end{center}

	The differences are found in L$\to$, R$\to$ and R$\forall$. Note how in particular the rules for R$\to$ and R$\forall$ allows only a single formula in the succedent.

	\pagebreak

	\begin{definition}
		A \textit{Derivation} is a rooted finite labelled tree, in which each leaf is labelled with one of the two axiom rules and each other label is derived from the labels of the successor nodes in accordance to the rules above. All formulas not in $\Gamma$ or $\Delta$ are the \textit{active} formulas of the rule. The label of the root is called \textit{head}. A formula that is the head of a derivation is said to be \textit{derivable}.
	\end{definition}

	The following is one of the fundamental results of first-order Logic:

	\begin{theorem}[Completeness theorem]
		A sequent $\Phi\Rightarrow \varphi$ is derivable iff $\Phi\models \varphi$.
	\end{theorem}
